Clustering time series based on probability distributions across temporal granularities

### 2021-11-01 main.Rmd
validation section done

To validate the clustering approaches, we create data designs that replicate prototype behaviors that might be seen in electricity data contexts. We spiked several attributes in the data to see where one method works better than the other and where they might give us the same outcome or the effect of missing data <!---and trends--> on the proposed methods. Three circular granularities $g1$, $g2$ and $g3$ are considered with categories denoted by $\{g10,g11\}$, $\{g20, g21, g22\}$ and $\{g30, g31, g32, g33, g34\}$ and levels $n_{g_1}=2$, $n_{g_2}=3$ and $l_{g_3}=5$. These categories could be integers or some more meaningful labels. For example, the granularity "day-of-week" could be either represented by ${0, 1, 2, \dots, 6}$ or ${Mon, Tue, \dots, Sun}$. Here categories of $g1$, $g2$ and $g3$ are represented by  $\{0, 1\}$, $\{0, 1, 2\}$ and $\{0, 1, 2, 3, 4\}$ respectively. A continuous measured variable $v$ of length $T$ indexed by $\{0, 1, \dots T-1\}$ is simulated such that it follows the structure across $g1$, $g2$ and $g3$. We constructed independent replications of all data designs $R = 25, 250, 500$ to investigate if our proposed clustering method can discover distinct designs in small, medium, and big number of series. All designs employ $T=300, 1000, 5000$ sample sizes to evaluate small, medium, and large sized series. Variations in method performance may be due to different jumps between categories. So a mean difference of $diff = 1, 2, 5$ is examined. The approaches' performance varies with the number of significant granularities. So all, few, and one major granularity scenarios are considered. Please see the Supplementary section (https://github.com/Sayani07/paper-gracsR) for the code and findings.

@@

Each category or combination of categories from $g1$, $g2$ and $g3$ are assumed to come from the same distribution, a subset of them from the same distribution, a subset of them from separate distributions, or all from different distributions, resulting in various data designs. As the methods ignore the linear progression of time, there is little value in adding time dependency in the data generating process. The data type is set to be "continuous," and the setup is assumed to be Gaussian. When the distribution of a granularity is "fixed", it means distributions across categories do not vary and are considered to be from N (0,1). The mean of different categories are altered in the "varying" designs, leading to varying distributions across categories.

@@

\noindent Consider the instance where $g1$, $g2$, and $g3$ all contribute to design distinction. Meaning that at least one of the designs to be categorized will have significantly different patterns for each granularity. In Table \ref{tab:tab-dist-design} (top), we explore various distributions across categories (as shown in Table Table \ref{tab:tab-dist-design} (bottom). Figure \ref{fig:plot-3gran-new} shows the simulated variable's linear and cyclic representations for each of these five designs. The structural difference in the time series variable is impossible to discern from the linear view. The shift in structure may be seen clearly in the distribution of cyclic granularities (Figure \ref{fig:plot-3gran-new} (right)). The following scenarios use solely graphical displays across cyclic granularities to highlight distributional differences in categories.

@@

\noindent This is the case where one granularity will remain the same across all designs. We consider the case where the distribution of $v$ would vary across levels of $g2$ for all designs, across levels of $g3$ for few designs and $g1$ does not change across designs. The proposed design is shown in Figure \ref{fig:gran2and1-clubbed}(left).

_(c) One signifiant granularity_

@@

```{r generate-design-3change}

@@

```{r plot-3gran-new, out.width="100%", fig.cap="The linear (left) and cyclic (right) representation of the measured variable is shown. In this scenario, all of $g1$, $g2$ and $g3$ changes across at least one design. Also, it is not possible to comprehend these patterns across cyclic granularities or group similar series just by looking at the linear plots."}

@@

```{r gran2and1-clubbed, fig.cap=" For the left scenario $g1$, $g2$ would change across atleast one design but $g3$ change remains same across all design. For the right one, only $g3$ changes across different designs."}
#gran2_change  + gran1_change + plot_layout(widths = c(1,1))
#ggpubr::ggarrange(gran1_change, gran2_change, labels = c("a", "b"))
```

@@

The proposed methods could be extended when two granularities of interest interact and we want to group subjects based on the interaction of the two granularities. Consider a group that has a different weekday and weekend behavior in the summer but not in the winter. This type of combined behaviour across granularities  can be discovered by evaluating the distribution across combinations of categories for different interacting granularities (wknd-wday and month-of-year in this example). As a result, in this scenario, we analyse a combination of categories generated from different distributions. Consider a case in which there are only two interacting granularities of interest, $g1$ and $g2$. In contrast to the previous situation, when we could study distributions across $n_{g_1} + n_{g_2} = 5$ separate categories, with interaction, we must evaluate the distribution of the $n_{g_1}*n_{g_2}=6$ combination of categories. Consider the $4$ designs in Figure \ref{fig:interaction-gran}, where various distributions are assumed for different combinations of categories, resulting in different designs. Design-1 exhibits no change in distributions across $g1$ or $g2$, whereas Design-2 and Design-3 alter across only $g1$ and $g2$, respectively. Design-4 varies across both $g1$ and $g2$ categories. Design-3 and Design-4 appear similar based on their relative difference across consecutive categories, however Design-4 also changes across facets, unlike Design-3, which has all facets look the same.

```{r interaction-gran, fig.cap = "Design-1 (a) has no change in distributions across different categories of $g1$ or $g2$, while Design-2 (b) and Design-3 (c) change across only $g1$ and $g2$ respectively. Design-4 (d) changes across categories of both $g1$ and $g2$."}

@@

- In Figure \ref{fig:mds-plot-validation}, we tried to see how separated our clusters are. We observe that in all scenarios and for different mean differences, cluster is separated. However, the separation increases with increase in mean differences across scenarios. This is intuitive because with increasing difference between categories, it gets easier for the methods to correctly distinguish the designs. Results corresponding to Scenario (a) is shown here.

- Figure \ref{fig:parcoord-sim} depicts a parallel coordinate plot with the vertical bar showing total inter-cluster distances with regard to granularities $g1$, $g2$, and $g3$. For all simulation settings and scenarios, values are represented as a sequence of lines connected across each axis. One line in the figure, for example, shows the inter-cluster distances for one simulation scenario. The lines are not coloured by group since the purpose is to highlight the contribution of the factors to categorization rather than class separation. The first plot shows that no variable stands out in the clustering, but the following two designs show that {g1} and {g1, g2} have very low inter cluster distances, meaning that they did not contribute to the clustering. It is worth noting that these facts correspond to our original assumptions when developing the scenarios, which incorporate distributional differences over three (a), two (b), and one (c) significant granularities. Hence, Figure \ref{fig:parcoord-sim} (a), (b), and (c) validate the construction of scenarios (a), (b), and (c) respectively.

 - The js-robust and wpd methods perform worse for $nT=300$, then improve for higher $nT$ evaluated in the study. Although, with the type of residential load data sets, a complete year of load is the minimum requirement to capture predicted differences in winter and summer profiles, for example. Even if the data is only available for a month, $nT$ is expected to be at least $1000$ with half-hourly data. Hence, practically this is not a challenge as long as the performance is promising for higher $nT=300$.

- For smaller difference between categories, it is seen that method js-nqt would preform better js-robust in our study sample. In our study sample, method js-nqt outperforms method js-robust for smaller differences between categories. More testing, however, is required to corroborate this.

For more detailed plots and tables, please refer to the supplementary paper.

```{r  parcoord-sim, fig.cap="The parallel coordinate plot illustrates the total inter-cluster distances for granularities $g1$, $g2$, and $g3$. One line in the figure depicts the inter-cluster distances for a single simulation scenario. While the first plot indicates that no variable stands out during clustering, the next two designs demonstrate that {g1} and {g1, g2} have extremely low inter-cluster distances, indicating that they did not contribute to clustering. It is worth emphasising that these facts are consistent with our initial assumptions when designing the scenarios. "}

@@

```{r mds-plot-validation, fig.cap = "Cases are projected through MDS and the first two dimensions are shown to display the relative position of clusters for different simulation parameters. The rows of the grid represent the different mean differences in designs and the columns represent different designs. It can be observed that clusters become more compact and separated for higher distributional differences between categories."}

@@

For clustering based on JS-based distances, we chose the optimal number of clusters using (@Hennig2014-ah) as $5$. The distribution of electricity demand for the selected $24$ customers across hod, moy and wdwn are shown in Figure \ref{fig:hod-ind-group} (d, e, f). Our methodology is useful for grouping similar distributions over hod and moy and they are placed closely for easy comparison. Of course, certain customers in each group have distributions that differ from other members in the same group. However, it appears that the aim of grouping comparable distributions over considered variables has been accomplished to some extent. Figure \ref{fig:combined-groups-js} shows the summarized distributions across $5$ groups and assists us in characterizing each cluster. Figure \ref{fig:combined-groups-js} shows Groups $2$ and $5$ show a stronger hod pattern with a typical morning and evening peak, whereas groups 1, 3, and 5 show a moy pattern  with higher usage in winter months. Differences in wknd-wday between groups are not discernible, implying that it may not be a relevant variable in distinguishing various clusters.

@@

```{r combined-groups-js, fig.cap = "The distribution of electricity demand for the clusters across hod (a), moy (b) and wkndwday (c). It seems like group 2 and 5 have a hod pattern across its members, while group 1, 3, 5 have a moy pattern. Wknd-wday variations across groups are not distinguishable, indicating that it is not a critical variable for clustering. It is helpful to compare the summarised distributions of groups to that of individuals to confirm that the most of individuals in the group have the same characterisation."}

@@

Things become far more complicated when we consider a larger data set with more uncertainty, as they do with any clustering problem. Summarizing distributions across clusters with varied or outlying customers can result in a shape that does not represent the group. Furthermore, combining hetergeneous customers may result in similar-looking final clusters that are not effective for visually differentiating them. It is also worth noting that the wknd-wday behavior in the given case does not characterize any cluster. This, however, will not be true for all of the customers in the data set. If more extensive prototype selection is used, resulting in more comprehensive prototypes in the data set, this method might be used to classify the entire data set into these prototype behaviors. However, the goal of this section was to have a few customers that have significant patterns over one or more cyclic granularities, apply our clustering methodology to cluster them, and demonstrate that the method produces useful clusters.

>>>>

To validate the clustering approaches, we spiked several attributes in the data to see where one method works better than the other and where they might give us the same outcome <!--or the effect of missing data <!---and trends--> on the proposed methods. Three circular granularities $g1$, $g2$ and $g3$ are considered with categories denoted by $\{g10,g11\}$, $\{g20, g21, g22\}$ and $\{g30, g31, g32, g33, g34\}$ and levels $n_{g_1}=2$, $n_{g_2}=3$ and $n_{g_3}=5$. These categories could be integers or some more meaningful labels. For example, the granularity "day-of-week" could be either represented by $\{0, 1, 2, \dots, 6\}$ or $\{Mon, Tue, \dots, Sun\}$. Here categories of $g1$, $g2$ and $g3$ are represented by  $\{0, 1\}$, $\{0, 1, 2\}$ and $\{0, 1, 2, 3, 4\}$ respectively. A continuous measured variable $v$ of length $T$ indexed by $\{0, 1, \dots T-1\}$ is simulated such that it follows the structure across $g1$, $g2$ and $g3$. We constructed independent replications of all data designs $R = \{25, 250, 500\}$ to investigate if our proposed clustering method can discover distinct designs in small, medium, and big number of series. All designs employ $T=\{300, 1000, 5000\}$ sample sizes to evaluate small, medium, and large sized series. Variations in method performance may be due to different jumps between categories. So a mean difference of $\mu = \{1, 2, 5\}$ is examined. The performance of the approaches varies with the number of granularities which has interesting patterns across its categories. So scenarios with all, few, and one significant granularities are considered.

@@

Each category or combination of categories from $g1$, $g2$ and $g3$ are assumed to come from the same distribution, a subset of them from the same distribution, a subset of them from separate distributions, or all from different distributions, resulting in various data designs. As the methods ignore the linear progression of time, there is little value in adding time dependency to the data generating process. The data type is set to be "continuous," and the setup is assumed to be Gaussian. When the distribution of a granularity is "fixed", it means distributions across categories do not vary and are considered to be from N (0,1). $\mu$ alters in the "varying" designs, leading to varying distributions across categories.

@@

\noindent Consider the instance where $g1$, $g2$, and $g3$ all contribute to design distinction. This means that at least one of the designs to be categorized will have significantly different patterns for each granularity. In Table \ref{tab:tab-dist-design} (top), we explore various distributions across categories (as shown in Table \ref{tab:tab-dist-design} (bottom). Figure \ref{fig:plot-3gran-new} shows the simulated variable's linear (left) and cyclic (right) representations for each of these five designs. The structural difference in the time series variable is impossible to discern from the linear view, with all of them looking very similar. The shift in structure may be seen clearly in the distribution of cyclic granularities. The following scenarios use solely graphical displays across cyclic granularities to highlight distributional differences in categories.

@@

\noindent This is the case where one granularity will remain the same across all designs. We consider the case where the distribution of $v$ varies across $g2$ levels for all designs, across $g3$ levels for a few designs, and $g1$ does not vary across designs. The proposed design is shown in Figure \ref{fig:gran2and1-clubbed}(b).

_Scenario (c): One signifiant granularity_

@@

```{r generate-D3change}

@@

```{r plot-3gran-new, out.width="100%", fig.cap="The linear (left) and cyclic (right) representation of the simulated variable is shown. Each row represents a design in Scenario (a). In this scenario, all of $g1$, $g2$ and $g3$ changes across at least one design. Also, it is not possible to comprehend these differences in patterns just by looking at or considering the linear representation."}

@@

```{r gran2and1-clubbed, fig.cap=" Plots (b) and (c) correspond to Design (b) and (c) respectively. In (b) $g2$, $g3$ changes across atleast one design but $g1$ remains constant. Only $g3$ changes across different designs in (c)."}
#gran2_change  + gran1_change + plot_layout(widths = c(1,1))
#ggpubr::ggarrange(gran1_change, gran2_change, labels = c("a", "b"))
```

@@

The proposed methods could be extended when two granularities of interest interact and we want to group subjects based on the interaction of the two granularities. Consider a group that has a different weekday and weekend behavior in the summer but not in the winter. This type of combined behaviour across granularities  can be discovered by evaluating the distribution across combinations of categories for different interacting granularities (Weekend/Weekday and month-of-year in this example). As a result, in this scenario, we analyse a combination of categories generated from different distributions. Consider a case in which there are only two interacting granularities of interest, $g1$ and $g2$. In contrast to the previous situation, when we could study distributions across $n_{g_1} + n_{g_2} = 5$ separate categories, with interaction, we must evaluate the distribution of the $n_{g_1}*n_{g_2}=6$ combination of categories. Consider the $4$ designs in Figure \ref{fig:interaction-gran}, where various distributions are assumed for different combinations of categories, resulting in different designs. Design $D1$ exhibits no change in distributions across $g1$ or $g2$, whereas  Designs $D2$ and $D3$ alter across only $g1$ and $g2$, respectively. D4 varies across both $g1$ and $g2$ categories. D3 and D4 appear similar based on their relative differences across consecutive categories, but $D4$ also changes across facets, unlike $D3$, which has all facets look the same.

```{r interaction-gran, fig.cap = "Distribution of the simulated variable across $g1$ conditional on $g2$ is shown through boxplots for 4 designs. D1 has no change in distributions across different categories of $g1$ or $g2$, while D2 and D3 change across only $g1$ and $g2$ respectively. D4 changes across categories of both $g1$ and $g2$."}

@@

- In Figure \ref{fig:mds-plot-validation}, we tried to see how separated our clusters are. We observe that in all scenarios and for different mean differences, cluster is separated. However, the separation increases with an increase in mean differences across scenarios. This is intuitive because, as the difference between categories increases, it gets easier for the methods to correctly distinguish the designs.

- Figure \ref{fig:parcoord-sim} depicts a parallel coordinate plot with the vertical bar showing total inter-cluster distances with regard to granularities $g1$, $g2$, and $g3$. For all simulation settings and scenarios, values are represented as a sequence of lines connected across each axis. One line in the figure, for example, shows the inter-cluster distances for one simulation setting. The lines are not coloured by group since the purpose is to highlight the contribution of the factors to categorization rather than class separation. The first plot shows that no variable stands out in the clustering, but the following two designs show that {g1} and {g1, g2} have very low inter cluster distances, meaning that they did not contribute to the clustering. It is worth noting that these facts correspond to our original assumptions when developing the scenarios, which incorporate distributional differences over three (a), two (b), and one (c) significant granularities. Hence, Figure \ref{fig:parcoord-sim} (a), (b), and (c) validate the construction of scenarios (a), (b), and (c) respectively.

 - The js-robust and wpd methods perform worse for $nT=300$, then improve for higher $nT$ evaluated in the study. Although, a complete year of data is the minimum requirement to capture distributional differences in winter and summer profiles, for example. Even if the data is only available for a month, $nT$ with half-hourly data is expected to be at least $1000$. As a result, as long as the performance is promising for higher $nT=300$, this is not a challenge.

- In our study sample, the method js-nqt outperforms the method js-robust for smaller differences between categories. More testing, however, is required to corroborate this.

For more detailed results, please refer to the supplementary paper. The code for creating these data designs and running the methodologies is available at (https://github.com/Sayani07/paper-gracsR/Validation).

```{r  parcoord-sim, fig.cap = "The parallel coordinate plot illustrates the total inter-cluster distances for granularities $g1$, $g2$, and $g3$. One line in the figure depicts the inter-cluster distances for a single simulation scenario. While the first plot indicates that no variable stands out during clustering, the next two designs demonstrate that {g1} and {g1, g2} have extremely low inter-cluster distances, indicating that they did not contribute to clustering. It is worth emphasising that these facts are consistent with our initial assumptions when designing the scenarios and (a), (b), and (c) correspond to Scenario (a), (b) and (c) respectively."}

@@

```{r mds-plot-validation, fig.cap = "Relative positions of clusters corresponding to different scenarios (columns) for different values of mean differences between categories (rows) are shown using the first two dimensions of MDS. It can be observed that clusters become more compact and separated for higher mean differences between categories across all designs. Between designs, separation is least prominent corresponding to scenario (c) where only granularity is responsible for the clusters."}

@@

For clustering based on JS-based distances, we chose the optimal number of clusters using (@Hennig2014-ah) as $5$. The distribution of electricity demand for the selected $24$ customers across hod, moy and wdwn are shown in Figure \ref{fig:hod-ind-group} (d, e, f). Our methodology is useful for grouping similar distributions over hod and moy and they are placed closely for easy comparison. Of course, certain customers in each group have distributions that differ from other members in the same group. However, it appears that the aim of grouping comparable distributions over considered variables has been accomplished to some extent. Figure \ref{fig:combined-groups-js} shows the summarized distributions across $5$ groups and assists us in characterizing each cluster. Figure \ref{fig:combined-groups-js} shows Groups $2$ and $5$ show a stronger hod pattern with a typical morning and evening peak, whereas groups 1, 3, and 5 show a moy pattern  with higher usage in winter months. Differences in Weekend/Weekday between groups are not discernible, implying that it may not be a relevant variable in distinguishing various clusters.

@@

```{r combined-groups-js, fig.cap = "The distribution of electricity demand for the clusters across hod (a), moy (b) and wkndwday (c). It seems like group 2 and 5 have a hod pattern across its members, while group 1, 3, 5 have a moy pattern. Weekend/Weekday variations across groups are not distinguishable, indicating that it is not a critical variable for clustering. It is helpful to compare the summarised distributions of groups to that of individuals to confirm that the most of individuals in the group have the same characterisation."}

@@

Things become far more complicated when we consider a larger data set with more uncertainty, as they do with any clustering problem. Summarizing distributions across clusters with varied or outlying customers can result in a shape that does not represent the group. Furthermore, combining hetergeneous customers may result in similar-looking final clusters that are not effective for visually differentiating them. It is also worth noting that the Weekend/Weekday behavior in the given case does not characterize any cluster. This, however, will not be true for all of the customers in the data set. If more extensive prototype selection is used, resulting in more comprehensive prototypes in the data set, this method might be used to classify the entire data set into these prototype behaviors. However, the goal of this section was to have a few customers that have significant patterns over one or more cyclic granularities, apply our clustering methodology to cluster them, and demonstrate that the method produces useful clusters.
